---
title: "Reproducible Research: Peer Assessment 1"
author: "Xiaoan Xie"
date: "October 25, 2016"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
Download the zip file, unzip it and read into R
```{r setoptions, echo=FALSE, cache=TRUE, results='markup'}
getwd()

url<-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"

if(!dir.exists(("./wk2"))) dir.create("./wk2")

if(!file.exists("./wk2/data.zip")) download.file(url, destfile = "./wk2/data.zip")

if(!file.exists("./wk2/activity.csv")) unzip("./wk2/data.zip", exdir="./wk2")

activity<-read.csv("./wk2/activity.csv")
```

## What is mean total number of steps taken per day?
Calculate totle steps each day and store them in table "steps_per_dat_raw", here "raw" indicates that missing values are kept.
```{r}
steps_per_day_raw <- as.data.frame.table(tapply(activity$steps, 
                                                  activity$date, sum, na.rm=FALSE))
names(steps_per_day_raw)<-c("date", "total_steps")
```

Plot histogram plot of daily total steps
```{r}
library("ggplot2")
qplot(total_steps/1000, data=steps_per_day_raw, xlab="Total Steps per Day (10^3)", main = "Daily Total Steps")
```

Calculate mean and median steps per day excluding NA.
```{r}
mean_step_per_day_raw<-mean(steps_per_day_raw$total_steps, na.rm = TRUE)
median_step_per_day_raw<-median(steps_per_day_raw$total_steps, na.rm = TRUE)
```
#### The average daily total steps is `r mean_step_per_day_raw`, the median daily total steps is `r median_step_per_day_raw`

## What is the average daily activity pattern?
Calculate average steps in each 5 minutes interval 
```{r}
steps_mean_5.min_raw <- as.data.frame.table(
    tapply(activity$steps, activity$interval, mean, na.rm = TRUE))
names(steps_mean_5.min_raw)<-c("time", "average_steps")
```

Plot to illustrate daily activity pattern
```{r}
library("ggplot2")
g<-ggplot(steps_mean_5.min_raw, aes(time, average_steps))
g+geom_point()+
    labs(x="Time")+
    labs(y="Average Steps")+
    labs(title="Average Steps in 5 min Interval Across All Day")+
    scale_x_discrete(breaks=c(0000, 0200, 0400, 0600, 0800, 1000, 1200,
                              1400, 1600, 1800, 2000, 2200, 2355))
```
                                                                
Find maxium average steps across all day in each 5 minutes intervlas and report the time of it
```{r}
max<-max(steps_mean_5.min_raw$average_steps)
max_step_5.min_index_raw<-(grep(max(steps_mean_5.min_raw$average_steps), steps_mean_5.min_raw$average_steps))
max_time_raw<-steps_mean_5.min_raw[max_step_5.min_index_raw,1]
```

#### The maxium average steps across all day is `r max` at time `r max_time_raw`

## Imputing missing values
Calculate total missing value in the data set
```{r}
total_NA<-sum(is.na(activity))

print(colSums(is.na(activity)), type="html")
```
#### There are `r total_NA` missing value in the data set in column "Steps"

Replace NA with the average steps in that time interval which are stored in "steps_mean_5.min_raw" dataset calculated previously
```{r}
activity_rmNA<-activity

activity_rmNA[is.na(activity_rmNA$steps), "steps"]<-
    steps_mean_5.min_raw[, "average_steps"]
## this code is amazing! it has the same result from codes below
# merge<-merge(activity_rmNA, steps_mean_5.min_raw, by.x="interval", by.y = "time")
# merge[is.na(merge$steps), "steps"]<-merge[is.na(merge$steps), "average_steps"]
# merge<-arrange(merge, date)
# merge<-merge[,1:3]
```

Calculate total steps per day after imputing NA
```{r}
steps_per_day_rmNA <- as.data.frame.table(tapply(activity_rmNA$steps, 
                                                activity_rmNA$date, sum))
names(steps_per_day_rmNA)<-c("date", "total_steps")
```

Calculate mean and median after imputing NA and repost
```{r}
mean_step_per_day_rmNA<-mean(steps_per_day_rmNA$total_steps)
median_step_per_day_rmNA<-median(steps_per_day_rmNA$total_steps)
```

#### The average daily steps after imputing NA is `r mean_step_per_day_rmNA`, while the median of them is `r median_step_per_day_rmNA`

Plot histogram chart of both data with or without NA to compare the impact of imputing NA
```{r}
g1<-qplot(total_steps/1000, data=steps_per_day_raw, 
          xlab=NULL, main="Total Steps per Day with NA")

g2<-qplot(total_steps/1000, data=steps_per_day_rmNA, 
          xlab="Total Steps per Day (10^3)", main = "Total Steps per Day with Imputed NA")

library("Rmisc")
multiplot(g1, g2)
# this function requires Rmisc package
```

#### Imputing NA by replacing them with the average steps in that time interval doesn't change the results much

## Are there differences in activity patterns between weekdays and weekends?

Create column to indicate whether that day is weekday or weekend.
```{r}
library("dplyr")

activity_rmNA$date<-as.Date(activity_rmNA$date)
activity_rmNA_WD<-mutate(activity_rmNA, 
                         weekdays=weekdays(activity_rmNA$date))
# above code generate detailed weekdays ("monday"...)
activity_rmNA_WD<-mutate(activity_rmNA_WD, 
                         weekday=ifelse(
                             activity_rmNA_WD$weekdays=="Saturday"|
                                 activity_rmNA_WD$weekdays=="Sunday",
                             "weekend", "weekday"))
activity_rmNA_WD$weekdays<-NULL
# combine weekdays into "Weekday" and "Weekend"
```

Calculate mean in each interval/weekday
```{r}
WDdata<-aggregate(activity_rmNA_WD$steps, by = list(activity_rmNA_WD$interval,
                                                  activity_rmNA_WD$weekday), 
                mean)
names(WDdata)<-c("interval", "weekday", "average_steps")
```

Plot to compare weekdays vs. weekends
```{r}
g3<-ggplot(WDdata, aes(interval, average_steps))
g3+geom_line(aes(color=weekday))+
    labs(x="Time")+
    labs(y="Average Steps")+
    labs(title="Average steps in Weekdays vs. Weekends")+
    theme_bw()
```    

#### Activity in weekends are lighter in the morning comaring with weekdays, and heavier in the day time, indicating a different lifing/working style in weekends
I didn't use panel plot here coz I want to show a direct comparison







